Appendix: New and simplified manual controls for projection and slice tours

### 2023-01-27 paper/appendix.Rmd
grammarly changes

The problem with the new simple method (Algorithm 1) is that the precise values for $V_m$ cannot be specified because the orthonormalisation will change them.

@@

If $d>2$ the process would be sequentially repeated in the same manner that Gram-Schmidt is applied sequentially to orthormalise the columns of a matrix. If $d=1$ no orthonormalisation is needed, and the projection vector would simply need to be normalized after each adjustment.

<!-- Need to think about how the values for selected variable are made exact, that is constrained orthonormalisation.

Need to think about checks, and error catching, maybe in implementation section. If selected variable has 0 coefficient, will that generate an orthonormalisation error?-->

@@

# Iterate and save images
# Unconstrained orthonormalisation
Anew <- A
for (i in 1:3) {
  Anew[vchange,] <- A[vchange,]*(1-i*eps)
  exact <- Anew[vchange,]
  Anew <- orthonormalize_rotated(Anew, vchange)
  plot_basis(Anew, vchange, vcol="black")
  points(exact[1], exact[2], pch=16, col="orange")
}

@@

For $d=2$ projections, the projection matrix is the sub-matrix $A$, of $O$ (an orthonormal basis for the p-dimensional space), formed by its first two columns as illustrated in Eq.~\ref{matrices}. Whereas orthonormality of the basis for the p-dimensional space is given by $e_i\cdot e_j=\delta_{ij},{i,j,=1,\cdots, p}$, orthonormality of the projection matrix is expressed as $\sum_{k=1}^p a_{ki} a_{kj}=\delta_{ij}, ~{i,j=1,2}, ~{k=1,\cdots,p}$. Movement of the cursor takes the two components ${a_{m1},a_{m2}}$ into a selected new value ${a^*_{m1},a^*_{m2}}$. Although the motion is constrained by $a^{*2}_{m1}+a^{*2}_{m2}\leq 1$, this is not sufficient to guarantee orthonormality of the new projection matrix. One possible algorithm to achieve this is
\begin{enumerate}
\item Cursor movement takes ${a_{m1},a_{m2}}\to {a^*_{m1},a^*_{m2}}$, called $V_m$ above.
\item The freedom to change the components $a_{m3}\cdots a_{mp}$ (the columns of $O$ not corresponding to the projection matrix $A$) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $a^*_{m3}=\sqrt{1-a^{*2}_{m1}-a^{*2}_{m1}},~a^*_{m,k>3}=0$
\item For other rows, $a_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalized), and then proceeding as usual $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc, resulting in $A^{**}$.
\end{enumerate}
\item this results in the orthonormal basis $O^*$ and a new projection matrix with $A^{**}_{m1}=a^*_{m1},~A^{**}_{m2}=a^*_{m2}$.
\end{enumerate}
The random completion of $O$ outside the projection matrix provides an exploration of dimensions orthogonal to the projection plane. However, it renders projections in a way that is not continuous  and may be distracting. This can be alleviated by replacing the random completion with a rule restricting the size of the jumps, for example in step (b) the values $a_{i\neq m,j\geq3}$ can be left unchanged before orthogonalisation.

@@

We could also take a similar approach without completing the basis, through rotation within the $d$-dimensional hyperplane defined by the new projection. In that case we would again start from $A^*$ as above, and then rotate the basis such that the first direction is aligned with the selected variable $m$. Applying Gram-Schmid within that basis definition will not alter the direction for the first vector, thus the direction of $m$ remains fixed to what was selected with the cursor. By adjusting the normalization procedure we can also ensure that the lenght remains as selcted manually by the user.

@@

```{r anchornav, fig.width=10, fig.height=3, out.width="100%", fig.cap="Visual guide for the slice center. Variable axes are displayed in polar coordinates, where the center corresponds to minimum value and outer end corresponds to maximum value. The position of the center corresponds to the dark polygon. If the center is at the center of the data, this will be displayed as a regular polygon (A), and plots B, C show it's position when moving the center along one axis."}
par(pty="s", mfrow=c(1,3))
plot(c(-1, 1), c(-1, 1), type="n", axes=FALSE, xlab = "", ylab = "", main="A. data center")
anchor = data.frame(matrix(rep(0.5, 6), ncol=6))
tourr:::draw_slice_center(
   anchor, 
   rng = matrix(rep(c(0,1), 6), ncol=6, byrow=FALSE),
   limits = 1, anchor_nav = "center")
plot(c(-1, 1), c(-1, 1), type="n", axes=FALSE, xlab = "", ylab = "", main="B. center moved towards max value of X3")
anchor[3] <- 0.75
tourr:::draw_slice_center(
   anchor, 
   rng = matrix(rep(c(0,1), 6), ncol=6, byrow=FALSE),
   limits = 1, anchor_nav = "center")
plot(c(-1, 1), c(-1, 1), type="n", axes=FALSE, xlab = "", ylab = "", main="C. center moved towards min value of X3")
anchor[3] <- 0.25
tourr:::draw_slice_center(
   anchor, 
   rng = matrix(rep(c(0,1), 6), ncol=6, byrow=FALSE),
   limits = 1, anchor_nav = "center")
```

@@

@@

Finally `VisualiseSliceDynamic` allows to discern which points exist inside the slice and outside it. This function differs from `SliceDynamic` in that it looks at only one data set but displays both the points inside and outside the slice. The display panel allows the user to specify the size of both sets of points (inside and outside) along with the slice height, center point and zoom level. The function and its arguments are given below:

@@

- `data`: A data matrix containing only one group. In this case the data matrix should not contain column labels.
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `height`: The initial height of the slice.
- `heightRange`: The range of slice heights to be explored.

>>>>

The problem with the new simple method (Algorithm 1) is that the precise values for $V_m$ cannot be specified because the orthonormalization will change them.

@@

If $d>2$ the process would be sequentially repeated in the same manner that Gram-Schmidt is applied sequentially to orthormalise the columns of a matrix. If $d=1$ no orthonormalization is needed, and the projection vector would simply need to be normalized after each adjustment.

<!-- Need to think about how the values for the selected variable are made exact, that is constrained orthonormalization.

Need to think about checks, and error catching, maybe in the implementation section. If the selected variable has a 0 coefficient, will that generate an orthonormalization error?-->

@@

# Iterate and save images
# Unconstrained orthonormalization
Anew <- A
for (i in 1:3) {
  Anew[vchange,] <- A[vchange,]*(1-i*eps)
  exact <- Anew[vchange,]
  Anew <- orthonormalize_rotated(Anew, vchange)
  plot_basis(Anew, vchange, vcol="black")
  points(exact[1], exact[2], pch=16, col="orange")
  text(Anew[vchange, 1], Anew[vchange, 2], expression(V[m]), pos=3, col="orange")

}

@@

For $d=2$ projections, the projection matrix is the sub-matrix $A$, of $O$ (an orthonormal basis for the p-dimensional space), formed by its first two columns as illustrated in Eq.~\ref{matrices}. Whereas orthonormality of the basis for the p-dimensional space is given by $e_i\cdot e_j=\delta_{ij},{i,j,=1,\cdots, p}$, orthonormality of the projection matrix is expressed as $\sum_{k=1}^p a_{ki} a_{kj}=\delta_{ij}, ~{i,j=1,2}, ~{k=1,\cdots,p}$. Movement of the cursor takes the two components ${a_{m1},a_{m2}}$ into a selected new value ${a^*_{m1},a^*_{m2}}$. Although the motion is constrained by $a^{*2}_{m1}+a^{*2}_{m2}\leq 1$, this is not sufficient to guarantee the orthonormality of the new projection matrix. One possible algorithm to achieve this is
\begin{enumerate}
\item Cursor movement takes ${a_{m1},a_{m2}}\to {a^*_{m1},a^*_{m2}}$, called $V_m$ above.
\item The freedom to change the components $a_{m3}\cdots a_{mp}$ (the columns of $O$ not corresponding to the projection matrix $A$) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $a^*_{m3}=\sqrt{1-a^{*2}_{m1}-a^{*2}_{m1}},~a^*_{m,k>3}=0$
\item For other rows, $a_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalized), and then $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc, resulting in $A^{**}$.
\end{enumerate}
\item this results in the orthonormal basis $O^*$ and a new projection matrix with $A^{**}_{m1}=a^*_{m1},~A^{**}_{m2}=a^*_{m2}$.
\end{enumerate}
The random completion of $O$ outside the projection matrix provides an exploration of dimensions orthogonal to the projection plane. However, it renders projections in a way that is not continuous and may be distracting. This can be alleviated by replacing the random completion with a rule restricting the size of the jumps, for example in step (b) the values $a_{i\neq m,j\geq3}$ can be left unchanged before orthogonalization.

@@

We could also take a similar approach without completing the basis, through rotation within the $d$-dimensional hyperplane defined by the new projection. In that case, we would again start from $A^*$ as above, and then rotate the basis such that the first direction is aligned with the selected variable $m$. Applying Gram-Schmid within that basis definition will not alter the direction for the first vector, thus the direction of $m$ remains fixed to what was selected with the cursor. By adjusting the normalization procedure we can also ensure that the length remains as selected manually by the user.

@@

```{r anchornav, fig.width=10, fig.height=3, out.width="100%", fig.cap="Visual guide for the slice center. Variable axes are displayed in polar coordinates, where the center corresponds to the minimum value and the outer end corresponds to the maximum value. The position of the center corresponds to the dark polygon. If the center is at the center of the data, this will be displayed as a regular polygon (A), and plots B and C show its position when moving the center along one axis."}
par(pty="s", mfrow=c(1,3))
plot(c(-1, 1), c(-1, 1), type="n", axes=FALSE, xlab = "", ylab = "", main="A. data center")
anchor = data.frame(matrix(rep(0.5, 6), ncol=6))
tourr:::draw_slice_center(
   anchor, 
   rng = matrix(rep(c(0,1), 6), ncol=6, byrow=FALSE),
   limits = 1, anchor_nav = "center")
plot(c(-1, 1), c(-1, 1), type="n", axes=FALSE, xlab = "", ylab = "", main="B. center moved towards max value of X3")
anchor[3] <- 0.75
tourr:::draw_slice_center(
   anchor, 
   rng = matrix(rep(c(0,1), 6), ncol=6, byrow=FALSE),
   limits = 1, anchor_nav = "center")
plot(c(-1, 1), c(-1, 1), type="n", axes=FALSE, xlab = "", ylab = "", main="C. center moved towards min value of X3")
anchor[3] <- 0.25
tourr:::draw_slice_center(
   anchor, 
   rng = matrix(rep(c(0,1), 6), ncol=6, byrow=FALSE),
   limits = 1, anchor_nav = "center")
```

@@

\textcolor{blue}{The slice center position is controlled via numerical input along each axis in the original command `SliceDynamic`. This provides numerical feedback on the position of the center. For easier navigation but without the numerical feedback, an implementation of the slice center control via a visual display like the one in Figure \ref{fig:anchornav} is provided in `SliceDynamicCC`.}

@@

Finally `VisualiseSliceDynamic` allows to discern which points exist inside the slice and outside it. This function differs from `SliceDynamic` in that it looks at only one data set but displays both the points inside and outside the slice. The display panel allows the user to specify the size of both sets of points (inside and outside) along with the slice height, center point, and zoom level. The function and its arguments are given below:

@@

- `data`: A data matrix containing only one group. In this case, the data matrix should not contain column labels.
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `height`: The initial height of the slice.
- `heightRange`: The range of slice heights to be explored.

### 2023-01-27 paper/appendix.tex
grammarly changes

The problem with the new simple method (Algorithm 1) is that the precise
values for \(V_m\) cannot be specified because the orthonormalisation
will change them.

@@

If \(d>2\) the process would be sequentially repeated in the same manner
that Gram-Schmidt is applied sequentially to orthormalise the columns of
a matrix. If \(d=1\) no orthonormalisation is needed, and the projection
vector would simply need to be normalized after each adjustment.

@@

For \(d=2\) projections, the projection matrix is the sub-matrix \(A\),
of \(O\) (an orthonormal basis for the p-dimensional space), formed by
its first two columns as illustrated in
Eq.\textasciitilde{}\ref{matrices}. Whereas orthonormality of the basis
for the p-dimensional space is given by
\(e_i\cdot e_j=\delta_{ij},{i,j,=1,\cdots, p}\), orthonormality of the
projection matrix is expressed as
\(\sum_{k=1}^p a_{ki} a_{kj}=\delta_{ij}, ~{i,j=1,2}, ~{k=1,\cdots,p}\).
Movement of the cursor takes the two components \({a_{m1},a_{m2}}\) into
a selected new value \({a^*_{m1},a^*_{m2}}\). Although the motion is
constrained by \(a^{*2}_{m1}+a^{*2}_{m2}\leq 1\), this is not sufficient
to guarantee orthonormality of the new projection matrix. One possible
algorithm to achieve this is

\begin{enumerate}
\item Cursor movement takes ${a_{m1},a_{m2}}\to {a^*_{m1},a^*_{m2}}$, called $V_m$ above.
\item The freedom to change the components $a_{m3}\cdots a_{mp}$ (the columns of $O$ not corresponding to the projection matrix $A$) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $a^*_{m3}=\sqrt{1-a^{*2}_{m1}-a^{*2}_{m1}},~a^*_{m,k>3}=0$
\item For other rows, $a_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalized), and then proceeding as usual $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc, resulting in $A^{**}$.
\end{enumerate}
\item this results in the orthonormal basis $O^*$ and a new projection matrix with $A^{**}_{m1}=a^*_{m1},~A^{**}_{m2}=a^*_{m2}$.
\end{enumerate}

The random completion of \(O\) outside the projection matrix provides an
exploration of dimensions orthogonal to the projection plane. However,
it renders projections in a way that is not continuous and may be
distracting. This can be alleviated by replacing the random completion
with a rule restricting the size of the jumps, for example in step (b)
the values \(a_{i\neq m,j\geq3}\) can be left unchanged before
orthogonalisation.

@@

We could also take a similar approach without completing the basis,
through rotation within the \(d\)-dimensional hyperplane defined by the
new projection. In that case we would again start from \(A^*\) as above,
and then rotate the basis such that the first direction is aligned with
the selected variable \(m\). Applying Gram-Schmid within that basis
definition will not alter the direction for the first vector, thus the
direction of \(m\) remains fixed to what was selected with the cursor.
By adjusting the normalization procedure we can also ensure that the
lenght remains as selcted manually by the user.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{appendix_files/figure-latex/anchornav-1} \caption{Visual guide for the slice center. Variable axes are displayed in polar coordinates, where the center corresponds to minimum value and outer end corresponds to maximum value. The position of the center corresponds to the dark polygon. If the center is at the center of the data, this will be displayed as a regular polygon (A), and plots B, C show it's position when moving the center along one axis.}\label{fig:anchornav}
\end{figure}

@@

Finally \texttt{VisualiseSliceDynamic} allows to discern which points
exist inside the slice and outside it. This function differs from
\texttt{SliceDynamic} in that it looks at only one data set but displays
both the points inside and outside the slice. The display panel allows
the user to specify the size of both sets of points (inside and outside)
along with the slice height, center point and zoom level. The function
and its arguments are given below:

@@

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix containing only one group. In this case
  the data matrix should not contain column labels.
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{height}: The initial height of the slice.
\item
  \texttt{heightRange}: The range of slice heights to be explored.
\end{itemize}

>>>>

The problem with the new simple method (Algorithm 1) is that the precise
values for \(V_m\) cannot be specified because the orthonormalization
will change them.

@@

If \(d>2\) the process would be sequentially repeated in the same manner
that Gram-Schmidt is applied sequentially to orthormalise the columns of
a matrix. If \(d=1\) no orthonormalization is needed, and the projection
vector would simply need to be normalized after each adjustment.

@@

For \(d=2\) projections, the projection matrix is the sub-matrix \(A\),
of \(O\) (an orthonormal basis for the p-dimensional space), formed by
its first two columns as illustrated in
Eq.\textasciitilde{}\ref{matrices}. Whereas orthonormality of the basis
for the p-dimensional space is given by
\(e_i\cdot e_j=\delta_{ij},{i,j,=1,\cdots, p}\), orthonormality of the
projection matrix is expressed as
\(\sum_{k=1}^p a_{ki} a_{kj}=\delta_{ij}, ~{i,j=1,2}, ~{k=1,\cdots,p}\).
Movement of the cursor takes the two components \({a_{m1},a_{m2}}\) into
a selected new value \({a^*_{m1},a^*_{m2}}\). Although the motion is
constrained by \(a^{*2}_{m1}+a^{*2}_{m2}\leq 1\), this is not sufficient
to guarantee the orthonormality of the new projection matrix. One
possible algorithm to achieve this is

\begin{enumerate}
\item Cursor movement takes ${a_{m1},a_{m2}}\to {a^*_{m1},a^*_{m2}}$, called $V_m$ above.
\item The freedom to change the components $a_{m3}\cdots a_{mp}$ (the columns of $O$ not corresponding to the projection matrix $A$) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $a^*_{m3}=\sqrt{1-a^{*2}_{m1}-a^{*2}_{m1}},~a^*_{m,k>3}=0$
\item For other rows, $a_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalized), and then $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc, resulting in $A^{**}$.
\end{enumerate}
\item this results in the orthonormal basis $O^*$ and a new projection matrix with $A^{**}_{m1}=a^*_{m1},~A^{**}_{m2}=a^*_{m2}$.
\end{enumerate}

The random completion of \(O\) outside the projection matrix provides an
exploration of dimensions orthogonal to the projection plane. However,
it renders projections in a way that is not continuous and may be
distracting. This can be alleviated by replacing the random completion
with a rule restricting the size of the jumps, for example in step (b)
the values \(a_{i\neq m,j\geq3}\) can be left unchanged before
orthogonalization.

@@

We could also take a similar approach without completing the basis,
through rotation within the \(d\)-dimensional hyperplane defined by the
new projection. In that case, we would again start from \(A^*\) as
above, and then rotate the basis such that the first direction is
aligned with the selected variable \(m\). Applying Gram-Schmid within
that basis definition will not alter the direction for the first vector,
thus the direction of \(m\) remains fixed to what was selected with the
cursor. By adjusting the normalization procedure we can also ensure that
the length remains as selected manually by the user.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{appendix_files/figure-latex/anchornav-1} \caption{Visual guide for the slice center. Variable axes are displayed in polar coordinates, where the center corresponds to the minimum value and the outer end corresponds to the maximum value. The position of the center corresponds to the dark polygon. If the center is at the center of the data, this will be displayed as a regular polygon (A), and plots B and C show its position when moving the center along one axis.}\label{fig:anchornav}
\end{figure}

@@

Finally \texttt{VisualiseSliceDynamic} allows to discern which points
exist inside the slice and outside it. This function differs from
\texttt{SliceDynamic} in that it looks at only one data set but displays
both the points inside and outside the slice. The display panel allows
the user to specify the size of both sets of points (inside and outside)
along with the slice height, center point, and zoom level. The function
and its arguments are given below:

@@

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix containing only one group. In this case,
  the data matrix should not contain column labels.
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{height}: The initial height of the slice.
\item
  \texttt{heightRange}: The range of slice heights to be explored.
\end{itemize}
